---
title: "Data Structures"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Contents

* [Vectors](#vectors)
    * [Atomic Vectors](#atomic-vectors)
    * [Lists](#lists)
* [Attributes](#attributes)
    * [Factors](#factors)
* [Matrices and Arrays](#matrices-and-arrays)
* [Data Frames](#data-frames)
    * [Creation](#creation)
    * [Testing and Coercion](#testing-and-coercion)
    * [Combining Data Frames](#combining-data-frames)
    * [Special Columns](#special-columns)
    

Fundamental data structures in R can be grouped in two ways:

  * whether the type is _homogeneous_ or _hetergeneous_, and 
  * what is the _dimensions_ of the type.
  
This can be shown in the following table:

| Dimension | Homogeneous | Heterogeneous |
| --------- | ----------- | ------------- |
| 1d        | Atmoic Vector | List        |
| 2d        | Matrix      | Data Frame    |
| nd        | Array       |               |

`str()` can show the details of an R object, just like:

```{r}
str(c(1:3))
```

or:

```{r}
str(list(name = 'Mike', age = 13))
```

## Vectors

Here, the idiom `vector` refers to 'general' vectors, including (atomic) vectors and lists (or generic vectors).

The common properties of the atomic vectors and lists are:

  * Type, use `typeof()` to get the type.
  * Length, use `length()` to get how many elements inside the object.
  * Attribute, use `attributes()` to get/set meta information on the object.
  
Note: `is.vector()` does not test if an object is a vector but a vector without attributes apart from `names`.

```{r}
person <- list(name = "Mike", age = 13)
attributes(person)
is.vector(person)
attributes(person) <- list(class_name = 'person')
is.vector(person)
```

To test whether an R object is a vector or not, `is.atomic(x) || is.list(x)` should be used.

```{r}
is.list(person)
```


### Atomic Vectors

Types of atomic vectors: 

    * logical
    * integer
    * double (called as `numeric` in R)
    * character
    * complex
    * raw

Atomic vectos are created with `c()`, which is short for 'combine'. 

Be noted that, as atmoic vectors are homogenous, `c()` does not support nesting. Nested `c()` will be flattened to get a simple atomic vector.

```{r}
dbl_var <- c(1, 2.5, 4.5)
# With the L suffix, you get an integer rather than a double
int_var <- c(1L, 6L, 10L)
# Use TRUE and FALSE (or T and F) to create logical vectors
log_var <- c(TRUE, FALSE, T, F)
chr_var <- c("these are", "some strings")
```

Note that `T` or `F` is not short for **values** `TRUE` or `FALSE`. In fact, `T` and `F` are two **variables** defined inside R, and you can redefine them.

```{r}
exists('T')
exists('TRUE')
T <- .05
isTRUE(T)
```

`NA`:  `NA` represents for missing values. In essence, `NA` is a logical value, and it can always be coerced to the right type. If you want to specify, you can use `NA_real_`, `NA_integer_`, or `NA_character_` instead.

#### Types and tests

* `typeof()`
* 'is*' function: `is.logical`, `is.integer`, `is.double`, `is.numeric`, `is.character`, or `is.atomic`

The difference `is.numeric` and `is.double`:

```{r}
v <- 5L
is.numeric(v)
is.double(v)
```

`is.numeric` is a general test of an object being interpretable as numbers. Both `integer` and `double` are numeric, though most often `numeric` refers to `double`.

#### Coercion

All elements in an atomic vectors share the same type. So if elements of different types are given, coercion occurs to push all elements to the most flexible type. Normally the order of flexibility is:

logical - integer - double - character

Coercion is automatically invoked. To avoid unintentional coercion, `as.*` functions can be used to invoke explicit coercion.

### Lists

Lists are used to build up many of the more complicated data structures in R.

Lists are created by `list()`, the elements of a list can be of any types, including list itself.

```{r}
x <- list('A', 1:5, c(TRUE, FALSE, TRUE), list(c('another', 'list'), 'end'))
str(x)
```

A list is also called recursive structure.

```{r}
is.recursive(x)
```

`c()` will combine several list into one.

```{r}
# A list with four elements
x <- c(list(1, 2), list(3, 4))
str(x)
# A list of two lists
x <- list(list(1, 2), list(3, 4))
str(x)
# Vectors are coerced to lists
x <- c(list(1, 2), c(3, 4))
str(x)
```

`typeof()` a list is a `list`.

`is.list()` tests whether an object is a list.

`as.list()` coerces an object to a list.

`unlist()` unlists an object to an atomic vector. Necessary coercion may occurs.
